Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-06 survey/survey.tex
add some more references.

\item
The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
  Also since the monoidal units can be freely introduced and eliminated, sometimes
  they are omitted.  However, as this is in fact dangerous, as explained by
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest.

@@

% \begin{verbatim}
% examples:
%   (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
%   conditionals
%   toffoli
%   fredkin
%   revisit adder example from Sec 2
%   perhaps a larger example as well (from RC 18 paper?)
% \end{verbatim}

@@

But why are these particular coherence laws? Are they all necessary?
Conversely are they, in some appropriate sense, sufficient? This is
the so-called \emph{coherence problem}. Mac Lane, in his farewell address
as President of the American Mathematical Society~\cite{MacLane1976} gives
a good introduction and overview of such problems.  A more modern
interpretation (which can nevertheless be read into Mac Lane's own
exposition) would read as follows: given a set of equalities on abstract
words, regarded as a rewrite system, and two means of rewriting a word
in that language to another, is there some suitable notion of canonical
form that expresses the essential uniqueness of the non-trivial
rewrites?  Note how this word-and-rewrite problem is essentially
independent of the eventual interpretation. But one must take some care,
as there are obvious degenerate cases (involving ``trivial'' equations
involving $0$ or $1$) which lead to non-uniqueness. The landmark
results, first by Kelly-Mac Lane~\cite{KELLY197197} for closed
symmetric monoidal categories, then (independently) Laplaza and
Kelly~\cite{laplaza72,kelly74} for symmetric Rig Categories, is
that indeed there are sound and complete coherence conditions that
insure that all the ``obvious'' equalities between different abstract
words in these systems give rise to commutative diagrams. The
``obvious'' equalities come from \emph{syzygies} or
\emph{critical pairs} of the system of equations.

@@

The next combinator, $\absorbr \Leftrightarrow \absorbl$, is
particularly fascinating: while it says something simple --- that the
two obvious ways of transforming $0 * 0$ into $0$, namely absorbing
either the left or right zero --- it implies something subtle.  A
straightforward proof of $\absorbl$ which proceeds by saying that
$0 * t$ cannot be inhabited because the first member of the pair
cannot, is not in fact equivalent to $\absorbr$ on $0 * 0$.  However,
if we instead define $\absorbl$ to ``transport'' the putative
impossible first member of the pair to its (equally impossible)
output, then these do form equivalent pairs.  The next few in
Fig.~\ref{figc} also express how $\absorbr$ and $\absorbl$ interact
with other combinators. As seen previously, all of these arise as
critical pairs. What is much more subtle here is that the types
involved often are asymmetric: they do not have the same occurrences
on the left and right. Such cases are particularly troublesome for
finding normal forms.

>>>>

\item
The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
  Also since the monoidal units can be freely introduced and eliminated, sometimes
  they are omitted.  However, as this is in fact dangerous, as explained
  by~\cite{BLUTE1996229}, we will err on the side of including them.

@@

@@

But why are these particular coherence laws? Are they all necessary?
Conversely are they, in some appropriate sense, sufficient? This is
the so-called \emph{coherence problem}. Mac Lane, in his farewell address
as President of the American Mathematical Society~\cite{MacLane1976} gives
a good introduction and overview of such problems.  A more modern
interpretation (which can nevertheless be read into Mac Lane's own
exposition) would read as follows: given a set of equalities on abstract
words, regarded as a rewrite system, and two means of rewriting a word
in that language to another, is there some suitable notion of canonical
form that expresses the essential uniqueness of the non-trivial
rewrites?  Note how this word-and-rewrite problem is essentially
independent of the eventual interpretation. But one must take some care,
as there are obvious degenerate cases (involving ``trivial'' equations
involving $0$ or $1$) which lead to non-uniqueness. The landmark
results, first by Kelly-Mac Lane~\cite{KELLY197197} for closed
symmetric monoidal categories, then (independently) Laplaza and
Kelly~\cite{laplaza72,kelly74} for symmetric Rig Categories, is
that indeed there are sound and complete coherence conditions that
insure that all the ``obvious'' equalities between different abstract
words in these systems give rise to commutative diagrams. The
``obvious'' equalities come from \emph{syzygies} or
\emph{critical pairs} of the system of equations.
The problem is far from trivial --- Fiore et al.~\cite{Fiore-2008}
document some publications where the coherence set is in
fact incorrect. They furthermore give a quite general algorithm
to derive such coherence conditions.

@@

The next combinator, $\absorbr \Leftrightarrow \absorbl$, is
particularly fascinating: while it says something simple --- that the
two obvious ways of transforming $0 * 0$ into $0$, namely absorbing
either the left or right zero --- it implies something subtle.  A
straightforward proof of $\absorbl$ which proceeds by saying that
$0 * t$ cannot be inhabited because the first member of the pair
cannot, is not in fact equivalent to $\absorbr$ on $0 * 0$.  However,
if we instead define $\absorbl$ to ``transport'' the putative
impossible first member of the pair to its (equally impossible)
output, then these do form equivalent pairs.  The next few in
Fig.~\ref{figc} also express how $\absorbr$ and $\absorbl$ interact
with other combinators. As seen previously, all of these arise as
critical pairs. What is much more subtle here is that the types
involved often are asymmetric: they do not have the same occurrences
on the left and right. Such cases are particularly troublesome for
finding normal forms. Laplaza~\cite{laplaza72} certainly comments on this,
but in mostly terse and technical terms. Blute et al.~\cite{BLUTE1996229}
offer much more intuitive explanations.
